Further Evidence of Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2022-03-17 src/ms.tex
adding CMD figure
paper edits

In addition to the original CKS catalog, we also cross-matched our sample with the catalogs of \citet{Brewer2018} and \citet{Martinez2019}, both of which presented spectroscopic parameters for CKS stars based on independent analysis of the same spectra. The \citet{Brewer2018} study, referred to here as SPOCS, also published elemental abundances and ages from isochrone fitting for the CKS sample. We additionally cross-matched the CKS catalog with the LAMOST--Kepler catalog \citep{Xiang2019} which is described further in \S\ref{subsec:lamost}. We compare the \teff--\prot distributions of the CKS sample using \teff and rotation periods from a variety of sources in Appendix~\ref{app:teffprot}.

\subsection{LAMOST--Kepler} \label{subsec:lamost}
The LAMOST--Kepler project derived homogeneous spectroscopic parameters from low-resolution ($R\sim$~1800) LAMOST DR5 spectra for approximately 40\% of the Kepler field \citep{Zong2018, Xiang2019}. The \citet{Xiang2019} catalog derived stellar parameters using the DD-Payne pipeline, which builds on the method of \citet{Ting2017b} by incorporating elements of the Cannon \citep{Ness2015} and uses the overlap with GALAH DR2 and APOGEE DR14 as training data. We cross-matched the LAMOST--Kepler stellar parameter catalog of \citet{Xiang2019} with the \citet{McQuillan2014} catalog, which published rotation periods for $>$34000 Kepler targets, as well as the \citet{Santos2021} catalog, which provides rotation periods for $>$55000 FGKM stars observed by Kepler.

@@

In the CKS sample, there appears to be a clustering of stars above the ridge with \teff~$>6100$~K (seen most clearly in the top panels of Figure~\ref{fig:models}). This cluster of points has a similar slope in the \teff--\prot plane as the long-period pile-up, and does not reside on the harmonic of the ridge line as one might expect if the periods were erroneously determined. After inspecting the light curves of stars on the ridge and above the ridge in the temperature range 6100~K~$<$~ \teff~$<6300$~K we did not notice any obvious trends or differences between the two samples. The periods of the stars in the cluster above the ridge appear to be just as reliable as those on the ridge. Interestingly, a similar clustering of points is not observed in the LAMOST--Kepler sample and is less pronounced or absent when substituting the CKS temperatures with \teff from either \citet{Brewer2018} or \citet{Martinez2019}, two studies that independently derived spectroscopic parameters from the CKS spectra. Further inspection of the stars in this cluster revealed that they have anomalously large \teff discrepancies between the CKS and SPOCS catalogs, such that the SPOCS temperatures shift the stars onto the long-period pile-up. We conclude that the temperatures for these stars in the CKS catalog are too high by $\gtrsim$~100~K.

@@

We compared the \teff--\prot distribution of the CKS and LAMOST--Kepler samples with the theoretical predictions of \citet{vanSaders2019}. Those authors presented forward modeling simulations of the observed Kepler \prot distribution including theoretical models of stellar angular momentum evolution (for both the standard spin-down and WMB scenarios), a galactic population model, and a prescribed observational selection function. The theoretically predicted \teff--\prot distributions of \citet{vanSaders2019} are shown in relation to the observations in Figure~\ref{fig:models}. Neither model satisfactorily matches the observations, though the WMB model more closely matches the long-period edge of F-type and early G-type stars. The specific WMB prescription of \citet{vanSaders2019} adopted a critical Rossby number of \rocrit~=~2.08, leading to a pile-up that is located at larger \prot values (at fixed \teff) when compared to the observations. Figure~\ref{fig:shifted} shows the same models in relation to the data with constant \teff offsets applied to the data, which are derived in \S\ref{subsec:rossby}. The \teff offsets lead to better agreement between the data and models, although the long-period pile-up in the \lamostmcq sample appears to overlay the models only for \teff~$\gtrsim 6000$~K, possibly due to strong systematics in the LAMOST \teff scale (see Appendix~\ref{app:teff}).

@@

While better agreement between the WMB model and observations might be achieved with stalling at a lower \rocrit, it is also possible that there are systematic offsets in the \teff scales between the observations and models used in \citet{vanSaders2019}, as well as differences in the computation of \taucz.  We also note that, while the models were computed using a simulated Kepler stellar population and selection function, the actual observed population and selection function of the LAMOST--Kepler catalog may be slightly different.

@@

\citet{Hall2021} determined rotation periods for 91 main-sequence, asteroseismic Kepler targets from rotational splitting of asteroseismic oscillation frequencies. We found that the distribution of the asteroseismic sample in the \teff--\prot plane approximately matches the pile-up we observe, although the \citet{Hall2021} sample appears shifted slightly to either higher \prot or higher \teff values relative to the ridge in the LAMOST--Kepler sample while such an offset is either absent or not as apparent relative to the CKS sample. Such an offset could be due to differing spectroscopic temperature scales between the two studies. To investigate the origin of this offset, we cross-matched the \citet{Hall2021} sample with the LAMOST DR5 catalog \citep{Xiang2019}. We found a root mean square (RMS) of the residuals between the \citet{Hall2021} and LAMOST DR5 \teff measurements of 55~K, with a median offset of 41~K, such that the LAMOST \teff scale is cooler, on average. Though this discrepancy is modest, adopting the LAMOST \teff scale appeared to resolve most of the offset between the pile-up as seen in the two samples.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{./figures/asteroseismic.pdf}
    \caption{Comparison of CKS (left) and LAMOST--Kepler samples (right) with the \citet{Hall2021} main-sequence asteroseismic sample in the \teff--\prot plane. Note, the CKS, LAMOST, and \citet{Hall2021} samples derive \teff from distinct pipelines. A constant offset of +41~K was applied to the LAMOST \teff to bring that sample onto the same scale as the asteroseismic sample (see Appendix~\ref{app:teff}). Using LAMOST \teff, where it exists, for the \citet{Hall2021} sample brings that sample into even closer agreement with the long-period pile-up in the LAMOST-McQuillan sample.}
    \label{fig:asteroseismic}
\end{figure}

@@

We tested the hypothesis that the long-period pile-up observed in the LAMOST--Kepler sample is due to WMB by fitting constant Rossby models to the \prot boundary. For a given \teff, this model predicts \prot as:

@@

To obtain a reasonable description of the long-period edge we computed the 90th percentile of \prot values in overlapping \teff bins with centers located every 20~K between 4000~K and 7000~K and half-widths of 100~K. We found it was also necessary to omit stars with Ro~$>$~5/3 in this computation to match the long-period edge. Since we observed a break in the 90th percentile \prot curve near \teff~=~4250~K, due to smaller sample sizes at low \teff, we restricted the analysis below to stars with \teff$>~4250$~K. To compute approximate uncertainties on the 90th percentile curve we performed 100 bootstrapped resamplings in each bin, leaving out 50\% of the data for each resampling. The 90th percentile curve and its uncertainty was then computed from the mean and standard deviation of the boostrapped values. We computed the 10th percentile curve and its uncertainty similarly, as an approximation to the lower boundary of the \teff-\prot plane. We show these curves in relation to the full LAMOST--Kepler sample and to constant Rossby models in Figure~\ref{fig:mcmc}.

@@

@@

In Figure~\ref{fig:teffscales} we compare temperatures between the LAMOST--Kepler catalog and temperatures from other surveys. We find that the LAMOST \teff scale is consistently cooler than other surveys by $\sim$20--80~K, with the exception of the \citet{McQuillan2014} study which sourced photometric \teff estimates from the KIC \citep{Brown2011}. A LAMOST \teff scale which is systematically cooler provides support to the notion that \rocrit determined from LAMOST temperatures will be systematically underestimated.

@@

\begin{acknowledgments}
We thank Rodrigo Luger for helpful discussions and assistance with the \texttt{showyourwork!} package, Adrian Price-Whelan for providing the APOGEE--Kepler cross-match catalog, and Karl Jaehnig for plotting advice. It is a pleasure to thank the Stars Group at the American Museum of Natural History, and the Astronomical Data Group at the Flatiron Institute for helpful discussions.

>>>>

In addition to the original CKS catalog, we also cross-matched our sample with the catalogs of \citet{Brewer2018} and \citet{Martinez2019}, both of which presented spectroscopic parameters for CKS stars based on independent analysis of the same spectra. The \citet{Brewer2018} study, referred to here as SPOCS, also published elemental abundances and ages from isochrone fitting for the CKS sample. We additionally cross-matched the CKS catalog with the \edit1{LAMOST DR5} catalog \citep{Xiang2019} which is described further in \S\ref{subsec:lamost}. We compare the \teff--\prot distributions of the CKS sample using \teff and rotation periods from a variety of sources in Appendix~\ref{app:teffprot}.

\subsection{\edit1{LAMOST DR5}} \label{subsec:lamost}
The LAMOST project derived homogeneous spectroscopic parameters from low-resolution ($R\sim$~1800) LAMOST DR5 spectra for approximately 40\% of the Kepler field \citep{Zong2018, Xiang2019}. The \citet{Xiang2019} catalog derived stellar parameters using the DD-Payne pipeline, which builds on the method of \citet{Ting2017b} by incorporating elements of the Cannon \citep{Ness2015} and uses the overlap with GALAH DR2 and APOGEE DR14 as training data. We cross-matched the \edit1{LAMOST DR5} stellar parameter catalog of \citet{Xiang2019} with the \citet{McQuillan2014} catalog, which published rotation periods for $>$34000 Kepler targets, as well as the \citet{Santos2021} catalog, which provides rotation periods for $>$55000 FGKM stars observed by Kepler.

@@

In the CKS sample, there appears to be a clustering of stars above the ridge with \teff~$>6100$~K (seen most clearly in the top panels of Figure~\ref{fig:models}). This cluster of points has a similar slope in the \teff--\prot plane as the long-period pile-up, and does not reside on the harmonic of the ridge line as one might expect if the periods were erroneously determined. After inspecting the light curves of stars on the ridge and above the ridge in the temperature range 6100~K~$<$~ \teff~$<6300$~K we did not notice any obvious trends or differences between the two samples. The periods of the stars in the cluster above the ridge appear to be just as reliable as those on the ridge. Interestingly, a similar clustering of points is not observed in the \edit1{\lamostmcq} sample and is less pronounced or absent when substituting the CKS temperatures with \teff from either \citet{Brewer2018} or \citet{Martinez2019}, two studies that independently derived spectroscopic parameters from the CKS spectra. Further inspection of the stars in this cluster revealed that they have anomalously large \teff discrepancies between the CKS and SPOCS catalogs, such that the SPOCS temperatures shift the stars onto the long-period pile-up. We conclude that the temperatures for these stars in the CKS catalog are too high by $\gtrsim$~100~K.

@@

We compared the \teff--\prot distribution of the CKS and \edit1{\lamostmcq} samples with the theoretical predictions of \citet{vanSaders2019}. Those authors presented forward modeling simulations of the observed Kepler \prot distribution including theoretical models of stellar angular momentum evolution (for both the standard spin-down and WMB scenarios), a galactic population model, and a prescribed observational selection function. The theoretically predicted \teff--\prot distributions of \citet{vanSaders2019} are shown in relation to the observations in Figure~\ref{fig:models}. Neither model satisfactorily matches the observations, though the WMB model more closely matches the long-period edge of F-type and early G-type stars. The specific WMB prescription of \citet{vanSaders2019} adopted a critical Rossby number of \rocrit~=~2.08, leading to a pile-up that is located at larger \prot values (at fixed \teff) when compared to the observations. Figure~\ref{fig:shifted} shows the same models in relation to the data with constant \teff offsets applied to the data, which are derived in \S\ref{subsec:rossby}. The \teff offsets lead to better agreement between the data and models, although the long-period pile-up in the \lamostmcq sample appears to overlay the models only for \teff~$\gtrsim 6000$~K, possibly due to strong systematics in the LAMOST \teff scale (see Appendix~\ref{app:teff}).

@@

While better agreement between the WMB model and observations might be achieved with stalling at a lower \rocrit, it is also possible that there are systematic offsets in the \teff scales between the observations and models used in \citet{vanSaders2019}, as well as differences in the computation of \taucz.  We also note that, while the models were computed using a simulated Kepler stellar population and selection function, the actual observed population and selection function of the \edit1{\lamostmcq} may be slightly different.

@@

\citet{Hall2021} determined rotation periods for 91 main-sequence, asteroseismic Kepler targets from rotational splitting of asteroseismic oscillation frequencies. We found that the distribution of the asteroseismic sample in the \teff--\prot plane approximately matches the pile-up we observe, although the \citet{Hall2021} sample appears shifted slightly to either higher \prot or higher \teff values relative to the ridge in the \edit1{\lamostmcq} sample while such an offset is either absent or not as apparent relative to the CKS sample. Such an offset could be due to differing spectroscopic temperature scales between the two studies. To investigate the origin of this offset, we cross-matched the \citet{Hall2021} sample with the LAMOST DR5 catalog \citep{Xiang2019}. We found a root mean square (RMS) of the residuals between the \citet{Hall2021} and LAMOST DR5 \teff measurements of 55~K, with a median offset of 41~K, such that the LAMOST \teff scale is cooler, on average. Though this discrepancy is modest, adopting the LAMOST \teff scale appeared to resolve most of the offset between the pile-up as seen in the two samples.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{./figures/asteroseismic.pdf}
    \caption{Comparison of CKS (left) and \edit1{\lamostmcq} samples (right) with the \citet{Hall2021} main-sequence asteroseismic sample in the \teff--\prot plane. Note, the CKS, LAMOST, and \citet{Hall2021} samples derive \teff from distinct pipelines. A constant offset of +41~K was applied to the LAMOST \teff to bring that sample onto the same scale as the asteroseismic sample (see Appendix~\ref{app:teff}). Using LAMOST \teff, where it exists, for the \citet{Hall2021} sample brings that sample into even closer agreement with the long-period pile-up in the LAMOST-McQuillan sample.}
    \label{fig:asteroseismic}
\end{figure}

@@

We tested the hypothesis that the long-period pile-up observed in the \edit1{\lamostmcq} sample is due to WMB by fitting constant Rossby models to the \prot boundary. For a given \teff, this model predicts \prot as:

@@

To obtain a reasonable description of the long-period edge we computed the 90th percentile of \prot values in overlapping \teff bins with centers located every 20~K between 4000~K and 7000~K and half-widths of 100~K. We found it was also necessary to omit stars with Ro~$>$~5/3 in this computation to match the long-period edge. Since we observed a break in the 90th percentile \prot curve near \teff~=~4250~K, due to smaller sample sizes at low \teff, we restricted the analysis below to stars with \teff$>~4250$~K. To compute approximate uncertainties on the 90th percentile curve we performed 100 bootstrapped resamplings in each bin, leaving out 50\% of the data for each resampling. The 90th percentile curve and its uncertainty was then computed from the mean and standard deviation of the boostrapped values. We computed the 10th percentile curve and its uncertainty similarly, as an approximation to the lower boundary of the \teff-\prot plane. We show these curves in relation to the full \edit1{\lamostmcq} sample and to constant Rossby models in Figure~\ref{fig:mcmc}.

@@

\edit1{To further assess the evolutionary state of stars on the long-period pile-up, we constructed a color-magnitude diagram (CMD) from the Gaia DR2 photometry and parallaxes for the Kepler field, the \citet{McQuillan2014} and \citet{Santos2021} rotation period catalogs, the \citet{Hall2021} asteroseismic sample, and the CKS long-period pile-up stars from this work (Figure~\ref{fig:cmd}). As expected, the stars with detected periods from rotational brightness modulations are almost exclusively solar-type and lower-mass stars on the main-sequence. The \citet{Santos2021} catalog contains more stars, in part due to the higher sensitivity to more slowly rotating, evolved stars, relative to \citet{McQuillan2014}. CKS stars on the long-period pile-up clearly occupy a well-defined region of the main-sequence, which overlaps well with the \citet{Hall2021} asteroseismic sample.}

\begin{figure}
    \includegraphics[width=\linewidth]{./figures/cmd.pdf}
    \caption{Color-magnitude diagram.}
    \label{fig:cmd}
\end{figure}

@@

In Figure~\ref{fig:teffscales} we compare temperatures between the \edit1{LAMOST DR5} catalog \citep{Xiang2019} and temperatures from other surveys. We find that the LAMOST \teff scale is consistently cooler than other surveys by $\sim$20--80~K, with the exception of the \citet{McQuillan2014} study which sourced photometric \teff estimates from the KIC \citep{Brown2011}. A LAMOST \teff scale which is systematically cooler provides support to the notion that \rocrit determined from LAMOST temperatures will be systematically underestimated.

@@

\begin{acknowledgments}
We thank the anonymous referee for a thoughtful review, Rodrigo Luger for helpful discussions and assistance with the \texttt{showyourwork!} package, Adrian Price-Whelan for providing the APOGEE--Kepler cross-match catalog, and Karl Jaehnig for plotting advice. It is a pleasure to thank the Stars Group at the American Museum of Natural History, and the Astronomical Data Group at the Flatiron Institute for helpful discussions.
